Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2013-02-24 documents/mixtures/response_to_referee.txt

   This paper gives a particularly lucid, concise and clear
   description of a relatively old idea. As such, it represents a
   significant contribution to the literature as a pedagogical
   reference. It should be published but see 4. below.

We agree with the referee that the idea of using Gaussians to fit
galaxies is old but we don't agree with any implication that the
method or applied math results shown are old.  All references we can
find in the literature show *how* to do mixture-of-Gaussian fitting in
real imaging, but don't give mixture-of-Gaussian approximations to
known profiles.  This paper, in that sense, is more of an "applied
math meets astronomy" paper than the previous work.  We will return to
this below, but we *did* make changes to the text in response to this
general concern of the referee: We added sentences in the introductory
paragraphs and at the end of the paper to emphasize the difficulty of
rendering the profiles without these approximations.

To be more specific: we added some sentences to emphasize that the
currently used methods for rendering profiles are effectively mixtures
of delta functions (the profile is sampled at a finite set of
locations) and rendering requires as many PSF renderings as there are
sample points.  This is a huge computation.

We also added an introductory paragraph that points out that the code
complexity required to do straight profile rendering (ie without the
MoGs) correctly is large and this leads to engineering overheads in
testing and correctness that go way beyond the simple numerical speed
losses.  We alsoo bring up the point that conventional numerical
integration methods effectively treat the profiles as very complex
mixtures of kernels.

@@

   Page 7, paragraph 2: The statement "These -- M^lux = 6 and M^luv =
   8 -- are good compromises between mixture complexity and quality of
   fit" warrants more explanation. Everytime I read about using MoGs
   for galaxy fitting, I have the same reaction: it needs a large
   number of highly redundant to model something a single exponential
   profile. Fans of the eigenimage decompositions are probably
   appalled!  Granted the mathematical simplicity of MoGs can make up
   for this by enabling faster fitting times, but it is definitely an
   un-appealing aspect of the method.  It would be helpful to comment
   on this to counter a reaction that I think may be shared by
   others. More importantly, I do not quite understand how you decided
   on M^lux = 6 and M^luv = 8. If I look at Table 2, the badness for
   lux has clearly converged and there is no gain in going higher than
   M=6. I cannot say the same thing for the luv MoG. Badness goes down
   by a factor of 16 going from M=6 to M=8 and then goes down by a
   factor of 5 going from M=8 to M=10. How can you claim that your
   badness has "converged" for M=8 then? A quantitative threshold and
   its rationale are clearly needed here to justify the "good
   compromises" here. Please be more quantitative or clarify if I
   misunderstood something.

[ response ; what changes ]

   Equation 19: I did not quite understand your distinction between
   the \Delta X vector here and the x vector in equation 2. Please
   clarify.

Agreed.  We added three sentences shortly after equation (19)
explaining explicitly why we use \Delta x and not x here.  It relates
to the fact that the PSF is only used as a convolution kernel, never
as a model in itself.

### 2013-03-10 documents/mixtures/response_to_referee.txt

   Page 6, paragraph 2: "K-L divergence" --> "Kullback-Leibler
   divergence"

fixed.

   Page 6 - Definition of "best": The authors should really have been
   "bolder" by insisting on the use of an information-theoretic
   measure here rather than the "traditional" chi-squared. The MoGs
   are a natural fit to an image entropy measure because maximizing
   entropy, say, would naturally lead to minimizing the number of
   components in a MoG. Chi-squared is also biased at low
   signal-to-noise ratios.  Please consider.

Chi-squared would definitely be biased if what we were fitting in
astronomical images were distributions of photon arrivals with no
background or read noise.  The presence of background (sky photons)
and read noise makes chi-squared a better statistic for optimization
than, say, the Poisson likelihood.  (That wouldn't be true if we were
fitting to *actual* images in which we know the sky level and read
noise, but we aren't.  Here we are just making *replacements* for the
standard profiles in a generic setting.)  All that said, the *precise*
measure you would use in *principle* depends (weakly) on the
properties of the data you have.  We added two sentences to this
effect in the relevant place (right after the K-L divergence mention).

As for the information-theoretic things mentioned by the referee: Yes,
in principle things like AIC and BIC and Bayes integral etc measures
for "how many components to use" would be very useful.  Unfortunately,
unlike with the decision about chi-squared vs Poisson likelihood, the
value of the AIC or BIC does depend very very sensitively on the noise
model (object, sky, and read noise levels) in every pixel of every
image.  That is, someone doing this correctly would use a higher M
expansion for the brighter galaxies and a lower M expansion for the
fainter galaxies.  That is very relevant and important.  We added text
about this where we give the M_exp=6 advice.  We note that although it
might make sense to go to large M from a data precision point of view,
it doesn't from a *model* precision point of view, since the models
don't fit the data at this level (there are spiral arms, deviations
from ellipsoidal symmetry and so on), as the referee evidently knows!
We added text about this too.

@@

   Page 6: Why is "badness" computed in a one-dimensional numerical
   integral? For the sake of computational speed? What happens when
   dealing with heavily blended objects? Where is the center for the
   1D integral then? Please explain. A general explanation of how
   object deblending is performed with MoGs would be welcome.

All we are trying to do in this paper is come up with *replacements*
for the exponential and deVaucouleurs profiles and variants.  We are
not actually fitting real galaxy images.  Indeed that requires
deblending and so on.  Literally, all we are doing is replacing the
exp{r^{1/n}) with a sum of Gaussians.  We are not fitting real scenes.

In the Tractor papers which are forthcoming, we *are* solving this
deblending problem, but that is a separate matter, involving image
likelihoods, noise models, and priors (and utilities, perhaps) on
parameters, so it is way beyond the scope of this paper.  This paper
is just about the profiles *themselves*, not any specific data
analysis problem that we can solve with them.

In this paper *all* we are doing is the "applied math" problem of
replacing the analytic profiles with MoG approximations.  As we note
in the introductory paragraphs, *any* use of deV and exp profiles
involves some approximation when convolving with a realistic PSF.  We
are just arguing for using MoG approximations instead of the
alternatives, most of which are finite, adaptive samplings, which are
more computationally expensive and less accurate.

@@

   Page 7, paragraph 2: The statement "These -- M^lux = 6 and M^luv =
   8 -- are good compromises between mixture complexity and quality of
   fit" warrants more explanation. Everytime I read about using MoGs
   for galaxy fitting, I have the same reaction: it needs a large
   number of highly redundant to model something a single exponential
   profile. Fans of the eigenimage decompositions are probably
   appalled!  Granted the mathematical simplicity of MoGs can make up
   for this by enabling faster fitting times, but it is definitely an
   un-appealing aspect of the method.  It would be helpful to comment
   on this to counter a reaction that I think may be shared by
   others. More importantly, I do not quite understand how you decided
   on M^lux = 6 and M^luv = 8. If I look at Table 2, the badness for
   lux has clearly converged and there is no gain in going higher than
   M=6. I cannot say the same thing for the luv MoG. Badness goes down
   by a factor of 16 going from M=6 to M=8 and then goes down by a
   factor of 5 going from M=8 to M=10. How can you claim that your
   badness has "converged" for M=8 then? A quantitative threshold and
   its rationale are clearly needed here to justify the "good
   compromises" here. Please be more quantitative or clarify if I
   misunderstood something.

We don't mean to claim that the badness has converged, and we modified
the text to say so.  We just mean that as a trade-off between code
speed and precision, in *practice* this is a good place to stop.
Again, this is just for *replacing* the analytic profiles, not for
general galaxy fitting.  For general galaxy fitting, we agree that
many more Gaussians can be required, especially for nearby
well-resolved galaxies.

@@

   Equation 19: I did not quite understand your distinction between
   the \Delta X vector here and the x vector in equation 2. Please
   clarify.

Agreed.  We added three sentences shortly after equation (19)
explaining explicitly why we use \Delta x and not x here.  It relates
to the fact that the PSF is only used as a convolution kernel, never
as a model in itself.

In addition to the changes given above, we also made the following small changes:

### 2013-03-10 documents/mixtures/mixture_models.tex
finished responding to referee issues

We seek the \emph{best} MoG approximations.  This necessitates
definition of the word ``best''.  If we think of the profiles as being
two-dimensional probability distribution functions (for, say, the
arrivals of photons), then one natural choice is the Kullback-Leibler
divergence or similar cross-entropy or information-theoretic measure.
However, in typical astronomical imaging, the galaxy is superimposed
on a substantial, flat sky level, and the noise in the data is close
to Gaussian.  This suggests more chi-squared-like objectives.  We
adopt the latter, in part because they are most appropriate for our
specific proposed application (modeling \sdss-like astronomical
imaging), but experiments we have performed suggest that
information-theoretic objectives also lead to good results.

@@

In our work on \project{The Tractor}, we use the $M^{\lux}=6$ lux and
the $M^{\luv}=8$ luv profiles.  Our advice to users would be to do the
same.  We use the lux and luv over the exp and dev partly because of
their better behaviors numerically, and partly because they and we are
both part of the \project{SDSS} tradition.  These---$M^{\lux}=6$ and
$M^{\luv}=8$---are good compromises between mixture complexity ($M$)
and quality of fit (badness).  Also, even the best-fitting late-type
and early-type galaxies deviate from exponential and de~Vaucouleurs
fits by more than do these high-quality MoG approximations; no
precision is lost.

@@

\acknowledgements It is a pleasure to thank
      Brendon Brewer (Auckland),
      Jim Bosch (Princeton),
  and Kevin Bundy (IPMU)
for valuable comments.  This work was supported in part by NASA (grant
NNX12AI50G) and the NSF (grant IIS-1124794).  All the results shown in
the figures and tables and more are available in machine-readable form
from author DWH upon request.

>>>>

We seek the \emph{best} MoG approximations.  This necessitates
definition of the word ``best''.  If we think of the profiles as being
two-dimensional probability distribution functions (for, say, the
arrivals of photons), then one natural choice is the Kullback-Leibler
divergence or similar cross-entropy or information-theoretic measure.
However, in typical astronomical imaging, the galaxy is superimposed
on a substantial, flat sky level, and the noise in the data is close
to Gaussian.  This suggests more chi-squared-like objectives.  We
adopt the latter, in part because they are most appropriate for our
specific proposed application (modeling \sdss-like astronomical
imaging), but experiments we have performed suggest that
information-theoretic objectives also lead to good results.  In
principle the objective that should be used to make these
approximations could be slightly different in different contexts.  In
practice, since the dependence of ``best objective'' on data
properties is weak and because the point here is to make generic
replacements for the analytic profiles, chi-squared is a good choice.

@@

In our work on \project{The Tractor}, we use the $M^{\lux}=6$ lux and
the $M^{\luv}=8$ luv profiles.  Our advice to users would be to do the
same.  We use the lux and luv over the exp and dev partly because of
their better behaviors numerically, and partly because they and we are
both part of the \project{SDSS} tradition.  We find that
these---$M^{\lux}=6$ and $M^{\luv}=8$---are good compromises between
mixture complexity ($M$) and quality of fit (badness) for our purposes
fitting \project{SDSS} data.  We expect this to be true for many
typical data sets.  In principle, $M$ should rise as the data quality
improves, because better data requires better models, and as $M$
increases, the approximations to the analytic profiles improve.  It is
important to remember, however, if you are tempted to go to very large
$M$, that this can be effort wasted: Even the best-fitting late-type
and early-type galaxies deviate from exponential and de~Vaucouleurs
fits by more than do these high-quality MoG approximations; no
precision is lost.

@@

\acknowledgements It is a pleasure to thank
      Brendon Brewer (Auckland),
      Jim Bosch (Princeton),
      Kevin Bundy (IPMU),
      and an anonymous referee
for valuable comments.  This work was supported in part by NASA (grant
NNX12AI50G) and the NSF (grant IIS-1124794).  All the results shown in
the figures and tables and more are available in machine-readable form
from author DWH upon request.
